---
title: "A First Quarto Document"
author: "Your Name"
date: today
format: html
---

## What This Is

This document mixes three things in one file: **prose** (this paragraph), **math** written
in LaTeX, and **code** that actually runs when the document is rendered. Nothing below was
copied and pasted from a separate R session; the numbers and the figure are produced by the
code chunks you can see.

We will simulate data from the simple linear model
$$
  y_i = \beta_0 + \beta_1 x_i + \varepsilon_i,
  \qquad \varepsilon_i \sim \mathcal{N}(0, \sigma^2),
$$
fit it by ordinary least squares, and report the result.

## Simulate and Fit

The `set.seed()` call makes this reproducible. Everyone who renders this document gets the
same "random" data, which is the whole point.

```{r}
#| label: simulate
#| message: false
set.seed(5014)

n <- 60
x <- runif(n, 0, 10)
y <- 2 + 0.5 * x + rnorm(n, sd = 1)

fit <- lm(y ~ x)
coef(fit)
```

The estimated intercept is `r round(coef(fit)[1], 2)` and the estimated slope is
`r round(coef(fit)[2], 2)`, against true values of 2 and 0.5. Those two numbers in the
previous sentence are inline code, not typed by hand. Change the seed or the sample size and
they update themselves when the document is re-rendered.

## A Figure

The chunk below is set to `echo: false`, so the plotting code is hidden and only the figure
appears. It also comes with a caption.

```{r}
#| label: fig-scatter
#| fig-cap: "Simulated data with the ordinary least-squares fit overlaid."
#| fig-width: 6
#| fig-height: 3.8
#| echo: false
plot(x, y,
     pch = 19, col = "steelblue",
     xlab = "x", ylab = "y")
abline(fit, lwd = 2)
```

## A Small Table

`summary()` on the fitted model gives the usual coefficient table:

```{r}
#| label: coef-table
summary(fit)$coefficients
```

If the data changed tomorrow, you would not touch a single number in the prose or redraw the
figure by hand. You would render the document again, and everything above would update at
once. That is reproducible reporting and is why your homework will now be in `.qmd` files.
